---
title: "Dynamic SEM on ESM Data"
subtitle: "Structural Equation Modeling Course Project - KU Leuven"
author: "MohammadHossein Manuel Haqiqatkhah - R0607671"
date: "2019-08-19"
bibliography: ref.bib
output: pdf_document
---


```{r setup, include = FALSE, echo = FALSE}
list.of.packages <- c("tidyverse",
                      "plyr",
                      "qgraph",
                      "dplyr",
                      "lavaan")
new.packages <-
  list.of.packages[!(list.of.packages %in% installed.packages()[, "Package"])]
if (length(new.packages)) {
  install.packages(new.packages, repos='http://cran.us.r-project.org')
}
tmp <- lapply(list.of.packages, require, character.only = TRUE)

rm(list.of.packages, new.packages, tmp)
knitr::opts_chunk$set(echo = FALSE)

```

```{r handy-functions}

pagebreak <- function(){
  if(knitr::is_latex_output())
    return("\\newpage")
  else
    return('<div style="page-break-before: always;" />')
}

```


```{r model-making}

make.factors <- function(items,
                         factor.name = "f",
                         add.factor.name = TRUE,
                         num.times = 2,
                         skip.first = 0,
                         substr.length = 3){
  factor.postfix <- ""

  if(add.factor.name) factor.postfix <- paste0(".",factor.name)
  loadings <- items %>%
    substring(skip.first + 1, skip.first + substr.length) %>%
    paste0(factor.postfix, "*")
  loadings[1] <- "1*"
  
  out <- c()  
  for(time in 1:num.times){
    items.here <- items %>% paste0(".",time)
    RHS.string <- paste(loadings, items.here, sep = "", collapse = " + ")
    factor.string <- paste(factor.name, ".", time, " =~ ",
                           RHS.string, sep = "")
    out <- out %>% c(factor.string)
  }
  out %>% return()
}

ar <- function(items, op = "~"){
  paste(paste0(items,".2"),op, paste0(items,".1"))
}

my.lav.model.maker <- function(inp.lagged,
                               add.resid.serial.AR = TRUE,
                               add.resid.serial.COV = FALSE,
                               add.posneg = TRUE,
                               add.posneg.AR = TRUE,
                               add.communal.VAR = TRUE,
                               add.factor.COV = TRUE,
                               communal.names = c("f","g")){
  
  # add = c("residAR", "factorVAR", "residCOR")
  
  model.strings <- c()
  
  # building lists of items
  items.pos <- inp.lagged$items.pos
  items.neg <- inp.lagged$items.neg
  items.all <- inp.lagged$items.all
  

  # AR/AC terms
  if(add.resid.serial.AR | add.resid.serial.COV){
    op <- " ~ "
    if(add.resid.serial.COV) op <- " ~~ "
    added.string <- items.all %>% ar(op)
    model.strings <- model.strings %>% c(added.string)
  }
  
  # factor terms
  if(add.posneg){
    pos.string <- items.pos %>% make.factors(factor.name = "pos")
    neg.string <- items.neg %>% make.factors(factor.name = "neg")
    model.strings <- model.strings %>% c(pos.string, neg.string)
  }
  
  for(f.name in communal.names){
    added.string <- items.all %>% make.factors(factor.name = f.name)
    model.strings <- model.strings %>% c(added.string)
  }
  model.strings %>% return()
}


my.lagger <- function(input,
                      items,
                      identifiers = NULL){
    
  cnames.1 <- items %>% paste0(".1")
  cnames.2 <- items %>% paste0(".2")

  d.items <- input %>%
    select(items)
  d.identifiers <- input %>%
    select(identifiers)
  
  d.lagged <- d.items[-1,] %>% 
    rbind(NA)

  d.out <- cbind(d.identifiers, d.items, d.lagged)
  colnames(d.out) <- c(identifiers, cnames.1, cnames.2)

  d.out %>% return()
}

```


# Introduction

Experience sampling method (ESM)

```{r data-prep}

# load data
data <- read.csv("https://raw.githubusercontent.com/psyguy/sem-course-project/master/data/ESMdata.csv")

# subset the needed data
d <- data %>% select(concentrat, dayno, beepno,
                     contains("mood_"))
items.p <- c("mood_relaxed",
             "mood_satisfi",
             "mood_enthus",
             "mood_cheerf",
             "mood_strong")

items.n <- c("mood_down",
             "mood_irritat",
             "mood_lonely",
             "mood_anxious",
             "mood_suspic",
             "mood_guilty",
             "mood_doubt")
identifiers <- c("concentrat",
                 "dayno",
                 "beepno")

f <- make.factors(items = items.p,
                  factor.name = "pa",
                  num.times = 2,
                  skip.first = 5)

d.lagged.1 <- d %>% my.lagger(items = items.p, identifiers = identifiers)

fit.0 <- sem(f, data=d.lagged.1, std.lv=TRUE)
```



```{r cars}
library(lavaan)


# model 1 - canonical mimic model (using the '~' regression operator)
model1 <- '
    f1 =~ y1 + y2 + y3
    f1 ~ x1 + x2 + x3
'

# model 2 - seemingly the same (using the '<~' composite operator)
model2 <- '
    f1 =~ y1 + y2 + y3
    # f1 <~ x1 + x2 + x3
'
# model 3 - only reflective
model3 <- '
    f1 =~ y1 + y2 + y3
    # f1 <~ x1 + x2 + x3
'
# model 4 - only formative
model4 <- '
    f1 ~ y1 + y2 + y3
    # f1 <~ x1 + x2 + x3
'

# run lavaan
fit1 <- sem(model1, data=data, std.lv=TRUE)
fit2 <- sem(model2, data=data, std.lv=TRUE)
fit3 <- sem(model3, data=data, std.lv=TRUE)
# fit4 <- sem(model4, data=data, std.lv=TRUE)

# test equality - only the operators are different
all.equal(parameterEstimates(fit1), parameterEstimates(fit2))
```





## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.



```{r writing-bib}
bib <- "



"

writeLines(bib, con = "ref.bib")
```


# Load cache from a latter code chunk

Sometimes we may want to insert an object value early in the document, when the object has not been created. [For example](http://stackoverflow.com/q/18581027/559676), you may want to mention a result in your abstract using an inline R expression, but the result is calculated later in the report.

One solution is to `save()` the workspace in a `.RData` file in the end of the document, and `load()` it in the beginning, e.g.

```{r}
x = 'NOT YET AVAILABLE'  # an object to be used
if (file.exists('everything.RData')) load('everything.RData')
```

In the end of the document, you save the workspace:

```{r eval=FALSE}
save.image('everything.RData')
```

Then you can use `` `r x` `` after `everything.RData` is loaded. When it is not available, you will see `x` as ``r x``.

The function `load_cache()` is an alternative solution, which allows you to load the value of an object from a specific code chunk, when the chunk has been cached.

For example, there is no object called `y` that has been created in this document yet, but we can still insert it here: ``r knitr::load_cache('test-a', 'y')``, as long as it will be created in the code chunk with the label `test-a` later.

```{r test-a, cache=TRUE}
y <- 2*pi
```

The first time you compile the document, you will see `y` is `NOT AVAILABLE`, but when you compile it for the second time, you will see its value ``r y``.

You do not have to specify the object name in `load_cache()`, in which case the database will just be loaded, and you can use any objects available in the database as if they had been computed by the code chunk later.

